High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-02-09 PNAS_manuscript/PNAS_manuscript.tex

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$82 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$127 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon could challenge the hypothesis of a young inner core. The multiple observed paleointensity transitions from weak to strong in the Paleozoic and the Proterozoic present challenges on the identification of the timing of inner core nucleation based on paleointensity records alone. 
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}. However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when the inner core unambiguously exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously settle the timing of the initiation of inner core nucleation based on such few records that span a relatively short time period.

@@

These features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Magnetic imaging also finds that orientations of magnetic needles within the Duluth plagioclase crystals depict the orientation of remanence magnetizations as is shown by the oblique acquired remanence to the direction of an applied IRM in Figure \ref{fig:Petro_QDM}F. Upon applying a second IRM of 300 mT in the orthogonal direction on the Duluth anorthosite, the applied field failed to rotate the remanence directions in plagioclase 1 but resulted in the reversal of remanence in plagioclase 2 which is still oblique to the applied field direction (Fig. \ref{fig:Petro_QDM}). In contrast, the remanence magnetizations imaged in the Beaver River anorthosite xenoliths align parallel to the applied field directions upon applying a set of orthogonal external fields (Fig. \ref{fig:Petro_QDM}). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex magma \cite{Miller1990a}.

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of individual specimen absolute paleointensity results (square symbols) from this study and their averages and standard deviations at site (black bars with grey uncertainty boxes) and locality level (orange dashed line and bar). All results are corrected for cooling rate with a factor of 0.74. As discussed in the text, the Beaver River anorthosite xenoliths likely record geomagnetic field intensities at similar but distinct times during the cooling of the Beaver River diabase. To compare these observations with more recent paleointensity variability, we took random draws of 3 values from the PADM2M model \cite{Ziegler2011a} for the VADM during the past 2 million years for a total of 10000 iterations and show the distribution of the calculated paleointensity average value of each draw at a latitude of 22\textdegree N. }}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within certain anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Strong remanence anisotropy associated with the igneous foliation developed within anorthosite from the Stillwater Complex has been shown to lead to significant overestimation or underestimation of paleointensity values depending on the relative orientations between the fabrics and an applied magnetic field \cite{Selkin2000a}. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have low gamma values ranging from 0.9\textdegree$\;$ to 11.9\textdegree, with a median value of 4.2\textdegree$\;$(Table S1). Because these anorthosite specimens were collected with various orientations with respect to the outcrops, the angle between the applied lab field direction during paleointensity experiments with respect to any fabrics within the anorthosites specimens are expected to cover a range of angles. In addition, these gamma values are similar to those of Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite xenolith bulk samples do not have significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts without deviating due to a fabric \cite{Zhang2021b}. In addition, no resolvable magnetic fabric was observed in the anorthosite xenoliths through petrographic and magnetic imaging (Fig. \ref{fig:Petro_QDM}).

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$127 ZAm$^2$. While the ca. 1140 Abitibi dikes paleointensity estimates do not go to values as low as the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in \cite{Kodama2019}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

Given that the multiple records of a weak field can not all be a minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby also explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions (Fig. \ref{fig:PINT_compilation}; \citealp{Bono2019a}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

@@

\subsection*{Paleointensity result selection}
The following criteria were used as quality filters on the paleointensity results: (1) a maximum angular deviation (MAD;  \citealp{Kirschvink1980a}) of $<$20\textdegree; (2) scatter parameter ($\beta$;  \citealp{Coe1978a}) values of $<$15$\%$; (3) a deviation angle (DANG;  \citealp{Tauxe2004a}) of $<$5\textdegree; (4) fraction of remanence fitted for paleointensity estimate (FRAC;  \cite{Shaar2013a}) $>$0.6; (5) scatter statistic (SCAT;  \citealp{Shaar2013a}) = TRUE; (6) a maximum magnetic moment difference between adjacent zero-field steps (GAP-Max;  \citealp{Shaar2013a}) $<$ 0.25; (7) number of pTRM checks $>$ 2; (8) and number of measurements used for paleointensity determination $\geq$ 4. The MAD measures the scatter about the best-fit line through the natural remanent magnetization (NRM) steps in the selected interval for which the intensity is defined. DANG, the deviation angle, is the angle between the best-fit direction that is free floating and the direction between the centre of mass of the data and the origin of the vector component diagram \cite{Tauxe2004a}. Both MAD and DANG assess the directional variation of the NRM, with MAD measuring the scatter in the NRM directions and DANG assessing whether the component is trending toward the origin of the Zijderveld plot. $\beta$ is the ``scatter" parameter of  \cite{Coe1978a} and is the ratio of the standard error of the slope of the best-fit line of the selected NRM and pTRM points on an NRM/TRM plot to the absolute value of the slope. FRAC is the fraction of the NRM that is used in the best-fit line \cite{Shaar2013a}. The FRAC value was chosen to preferentially select samples with dominantly single-slope NRM/TRM plots. GAP-Max is the maximum gap between two points on the NRM/TRM plot determined by vector arithmetic. SCAT is a Boolean operator which uses the error on the best-fit slope of the selected data on the NRM/TRM plot to determine if the data are overly scattered. The parameter is used to assess pTRM checks in addition to assessing the degree to which IZZI steps are zigzagged. $\beta$, FRAC, GAP-Max and SCAT are all statistics to assess the behavior of NRM/TRM plots. See the Standard Paleointensity Definitions ( \citealp{Paterson2014a}; \url{https://earthorg/PmagPy/SPD/home.html}) for more details. Data analysis was conducted using Thellier GUI \cite{Shaar2013a} within the PmagPy software package \cite{Tauxe2016a}.

>>>>

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$82 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$127 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core. The multiple observed paleointensity transitions from weak to strong in the Paleozoic and the Proterozoic present challenges on identifying the onset of inner core nucleation based on paleointensity records alone. 
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was soon followed by initial crystallization of the inner core \cite{Bono2019a}. However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when we known the inner core exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously settle the timing of the initiation of inner core nucleation based on the current few paleointensity records.

@@

These features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D, F, H). Magnetic imaging also finds that orientations of magnetic needles within the Duluth plagioclase crystals depict the orientation of remanence magnetizations as is shown by the oblique acquired remanence to the direction of a first applied isothermal remanent magnetization (IRM) in Figure \ref{fig:Petro_QDM}F. Upon applying a second IRM with a field of 300 mT in the orthogonal direction, remanent magnetization moments of the Duluth anorthosite failed to rotate in plagioclase 1 but reversed in plagioclase 2 but still oblique to the applied field direction (Fig. \ref{fig:Petro_QDM}F, H). In contrast, the remanence magnetizations imaged in the Beaver River anorthosite xenoliths align more parallel to the applied field directions upon applying a set of orthogonal external fields (Fig. \ref{fig:Petro_QDM}E, G). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex magma \cite{Miller1990a}.

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of individual specimen absolute paleointensity results (square symbols) from this study and their averages and standard deviations at site (black bars with grey uncertainty boxes) and locality level (orange dashed line and bar). All results are corrected for cooling rate with a factor of 0.74. The Beaver River anorthosite xenoliths likely record geomagnetic field intensities at similar but distinct times during cooling of the Beaver River diabase, capturing secular variations. To compare these variable observations with more recent paleointensity variability, we took random draws of 3 values from the PADM2M model \cite{Ziegler2011a} for the virtual axial dipole moment values during the past 2 million years for a total of 10000 iterations and show the distribution of average paleointensity values of each draw calculated for a latitude of 22\textdegree N. }}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within certain anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Strong remanence anisotropy associated with the igneous foliation developed within anorthosite from the Stillwater Complex has been shown to lead to significant overestimation or underestimation of paleointensity values depending on the relative orientations between the fabrics and an applied magnetic field \cite{Selkin2000a}. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have low gamma values ranging from 0.9\textdegree$\;$ to 11.9\textdegree, with a median value of 4.2\textdegree$\;$(Table S1). Because these anorthosite specimens were oriented at various directions with respect to the outcrops, the angle between the applied lab field direction during paleointensity experiments with respect to any fabrics within the anorthosites specimens are expected to cover a wide range of angles. These gamma values of the anorthosite xenolith bulk specimens are similar to those of Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the bulk Beaver River anorthosite xenoliths do not have significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts without deviating due to a fabric \cite{Zhang2021b}. In addition, no resolvable magnetic fabric was observed in the anorthosite xenoliths through petrographic and magnetic imaging (Fig. \ref{fig:Petro_QDM}).

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$127 ZAm$^2$. While the ca. 1140 Abitibi dikes paleointensity estimates do not go to values as low as the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in \cite{Kodama2019a}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

Given that the multiple records of a weak field cannot all be the minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby also explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions (Fig. \ref{fig:PINT_compilation}; \citealp{Bono2019a}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

@@

\subsection*{Paleointensity result selection}
The following criteria were used as quality filters on the paleointensity results: (1) a maximum angular deviation (MAD;  \citealp{Kirschvink1980a}) of $<$20\textdegree; (2) scatter parameter ($\beta$;  \citealp{Coe1978a}) values of $<$15$\%$; (3) a deviation angle (DANG;  \citealp{Tauxe2004a}) of $<$5\textdegree; (4) fraction of remanence fitted for paleointensity estimate (FRAC;  \cite{Shaar2013a}) $>$0.6; (5) scatter statistic (SCAT;  \citealp{Shaar2013a}) = TRUE; (6) a maximum magnetic moment difference between adjacent zero-field steps (GAP-Max;  \citealp{Shaar2013a}) $<$ 0.25; (7) number of pTRM checks $>$ 2; (8) and number of measurements used for paleointensity determination $\geq$ 4. The MAD measures the scatter about the best-fit line through the natural remanent magnetization (NRM) steps in the selected interval for which the intensity is defined. DANG, the deviation angle, is the angle between the best-fit direction that is free floating and the direction between the centre of mass of the data and the origin of the vector component diagram \cite{Tauxe2004a}. Both MAD and DANG assess the directional variation of the NRM, with MAD measuring the scatter in the NRM directions and DANG assessing whether the component is trending toward the origin of the Zijderveld plot. $\beta$ is the ``scatter" parameter of  \cite{Coe1978a} and is the ratio of the standard error of the slope of the best-fit line of the selected NRM and pTRM points on an NRM/TRM plot to the absolute value of the slope. FRAC is the fraction of the NRM that is used in the best-fit line \cite{Shaar2013a}. The FRAC value was chosen to preferentially select samples with dominantly single-slope NRM/TRM plots. GAP-Max is the maximum gap between two points on the NRM/TRM plot determined by vector arithmetic. SCAT is a Boolean operator which uses the error on the best-fit slope of the selected data on the NRM/TRM plot to determine if the data are overly scattered. The parameter is used to assess pTRM checks in addition to assessing the degree to which IZZI steps are zigzagged. $\beta$, FRAC, GAP-Max and SCAT are all statistics to assess the behavior of NRM/TRM plots. See the Standard Paleointensity Definitions (\citealp{Paterson2014a}; \url{https://earthorg/PmagPy/SPD/home.html}) for more details. Data analysis was conducted using Thellier GUI \cite{Shaar2013a} within the PmagPy software package \cite{Tauxe2016a}.
